Perturbation Theory in the Complex Plane: Exceptional Points and Where to Find Them

### 2020-11-30 Manuscript/EPAWTFT.tex
updated some figures

Using the UHF framework allows the singlet ground state wave function to mix with triplet wave functions, 
leading to spin contamination where the wave function is no longer an eigenfunction of the $\Hat{\cS}^2$ operator.
The link between slow UMP convergence and this spin-contamination was first systematically investigated
by Gill \etal\ using the minimal basis \ce{H2} model.\cite{Gill_1988}
In this work, the authors compared the UMP series with the exact RHF- and UHF-based FCI expansions
and identified that the slow UMP convergence arises from its failure to correctly predict the amplitude of the
low-lying double excitation.
This erroneous description of the double excitation amplitude has the same origin as the spin-contamination in the reference
UHF wave function, creating the first direct link between spin-contamination and slow UMP convergence.\cite{Gill_1988}
Lepetit \etal\ later analysed the difference between perturbation convergence using the unrestricted MP 
and EN partitionings. \cite{Lepetit_1988}
They argued that the slow UMP convergence for stretched molecules arises from 
(i) the fact that the MP denominator (see Eq.~\ref{eq:EMP2})
tends to a constant value instead of vanishing, and (ii) the slow convergence of contributions from the 
singly-excited configurations that strongly couple to the doubly-excited configurations and first
appear at fourth-order.\cite{Lepetit_1988}
Drawing these ideas together, we believe that slow UMP convergence occurs because the single excitations must focus on removing
spin-contamination from the reference wave function, limiting their ability to fine-tune the amplitudes of the higher 
excitations that capture the correlation energy.

@@

>>>>

Using the UHF framework allows the singlet ground state wave function to mix with triplet wave functions, 
leading to spin contamination where the wave function is no longer an eigenfunction of the $\Hat{\cS}^2$ operator.
The link between slow UMP convergence and this spin-contamination was first systematically investigated
by Gill \etal\ using the minimal basis \ce{H2} model.\cite{Gill_1988}
In this work, the authors compared the UMP series with the exact RHF- and UHF-based FCI expansions
and identified that the slow UMP convergence arises from its failure to correctly predict the amplitude of the
low-lying double excitation.
This erroneous description of the double excitation amplitude has the same origin as the spin-contamination in the reference
UHF wave function, creating the first direct link between spin-contamination and slow UMP convergence.\cite{Gill_1988}

% LEPETIT CHAT
Lepetit \etal\ later analysed the difference between perturbation convergence using the unrestricted MP 
and EN partitionings. \cite{Lepetit_1988}
They argued that the slow UMP convergence for stretched molecules arises from 
(i) the fact that the MP denominator (see Eq.~\ref{eq:EMP2})
tends to a constant value instead of vanishing, and (ii) the slow convergence of contributions from the 
singly-excited configurations that strongly couple to the doubly-excited configurations and first
appear at fourth-order.\cite{Lepetit_1988}
Drawing these ideas together, we believe that slow UMP convergence occurs because the single excitations must focus on removing
spin-contamination from the reference wave function, limiting their ability to fine-tune the amplitudes of the higher 
excitations that capture the correlation energy.

@@

%------------------------------------------------------------------%
% Figure on the UMP critical point
%------------------------------------------------------------------%
\begin{figure*}[t]
	\begin{subfigure}{0.49\textwidth}
	\includegraphics[height=0.65\textwidth]{ump_critical_point}
	\subcaption{\label{subfig:ump_cp}}
    \end{subfigure}
	\begin{subfigure}{0.49\textwidth}
	\includegraphics[height=0.65\textwidth]{ump_ep_to_cp}
	\subcaption{\label{subfig:ump_ep_to_cp}}
    \end{subfigure}
	\caption{%
    \hugh{Modelling the RMP critical point using the asymmetric Hubbard dimer.
        (\subref{subfig:rmp_cp}) Exact critical points with $t=0$ occur on the negative real $\lambda$ axis (dashed).
        (\subref{subfig:rmp_cp_surf}) Modelling a finite basis using $t=0.1$ yields complex-conjugate EPs close to the
        real axis, giving a sharp avoided crossing on the real axis (solid).
    }
	\label{fig:RMP_cp}}
\end{figure*}
%------------------------------------------------------------------%

% RELATIONSHIP BETWEEN QPT AND UMP
\hughDraft{%
The ground-state EP observed in the UMP series also has a small imaginary part and falls close to the real axis. 
As the correlation strength increases, this EP moves closer to the real axis and closer to the radius of convergence at 
$\lambda = 1$. So can we understand this using the arguments related to the critical point?
Closed-shell case:
The work by Stillinger builds an argument around the HF potential $v_{\text{HF}}$ which, by itself, 
is repulsive and concentrated around the 
occupied orbitals. As $\lambda$ becomes increasingly positive along the real axis, this HF potential 
becomes attractive for $\lambda > 1$.
However, the explicit two-electron becomes increasingly repulsive for larger positive $\lambda$, 
until eventually single electrons are successively expelled from the molecule.
Effect of symmetry-breaking:
Things become more subtle for the symmetry-broken case because we must now consider the $\alpha$ and $\beta$ HF potentials.
When UHF symmetry breaking occurs in the Hubbard dimer, with the $\alpha$ electron localising on the left site, 
the $\alpha$ HF potential
will then be a repulsive interaction localised around the $\beta$ electron, so on the right site. 
The same is true for the $\beta$ HF potential.
Therefore, as $\lambda$ becomes greater than 1 and the HF potentials become attractive, 
there is a driving force for the $\alpha$ and $\beta$.
electrons to swap sites. If both electrons swap at the same time, then we get a double excitation. Therefore, we expect an 
avoided crossing as $\lambda$ is increased beyond 1. 
The "sharpness" of this avoided crossing is controlled by the strength of the electron-electron repulsion... see below.
}

\hughDraft{%
For small $U/t$, the HF potentials will be weak and the electrons will slowly delocalise over 
both sites as $\lambda$ increases. The UHF reference itself already has some degree of delocalisation over 
both sites as we are only just beyond the CFP. This leads to a
"shallow" avoided crossing, and the corresponding EPs corresponding with have non-zero imaginary parts. 
At larger $U/t$, the HF potentials will becomes increasingly repulsive and the "swapping" driving force will become stronger. 
Furthermore, the strong on-site repulsion will make delocalisation over both sites increasingly unfavourable. 
We therefore see that the electrons swap sites almost instantaneously as $\lambda$ passes through 1, as this is the threshold point 
where the HF potential becomes attractive. This very rapid change in the nature of the state corresponds to a "sharp" avoided
crossing, with EPs close to the real axis.
Note that, although this appears to be an avoided crossing with the first-excited state, 
by the time we have reached $\lambda \approx 1$,
we are beyond the excited-state avoided crossing and thus the first-excited state qualitatively corresponds to a double
excitation from the reference. This matches our expectation of both electrons swapping sites.
Okay we can't actually do $U/t = \infty$, but we can make it very large. With such a strong attractive HF potential and such strong
on-site repulsion, the electrons swapping process will occur exactly at $\lambda=1$. In this limit, the change is so sudden that it
now corresponds to a QPT in the perturbation approximation, and again the EPs fall on the real axis.
}

\hughDraft{%
By applying the separation of the Hamiltionian proposed by Stillinger, we have been able to rationalise why the UMP ground-state
occurs so close to radius of convergence and increasingly close to the real axis for large $U/t$. The key feature here is the fact that the
HF potential is different for the $\alpha$ and $\beta$ electrons, and by extension the potential felt by 
an electron is not strictly localised around
that electrons orbitals. This is completely different to the symmetric structure in the closed-shell case. Because of these different 
potentials, there can be a large driving force for spatial rearrangement occurring as soon as the HF potential becomes positive 
(at $\lambda=1$). 
This sudden change in electron distribution is made more extreme if the HF potential and electron repulsion dominate over the 
one-electron terms, which also corresponds to the regime of strong wave function symmetry breaking. 
We have therefore provided a physical motivation behind why this EP can behave like a QPT for strong symmetry breaking. 
This supports the conclusions in Antoine's report that the symmetry-broken EP behaves as a class $\beta$ singularity.
}
